---
title: "Joe Musgrove: The Ear Controversy"
author: "Ryan Ramirez"
date: "2023-01-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Background


In the MLB, it is against the rules to use a foreign substance by pitchers. Only
rosin is allowed to maintain the integrity of the game. Illegal foreign substances
include pine tar, spider tack, or any sticky stuff. These are illegal because
previous research has found that spin rates increase dramatically when these
foreign substances are used by pitchers. In the recent years, the MLB has advocated
for pitchers to be checked before, during, and after the games for foreign substances.

One case in particular, is the controversy of Joe Musgrove's ears during the 2022
wild card series between the San Diego Padres and the New York Mets. In the
bottom of the 6th inning, Buck Showalter, manager of the Mets, talked to the 
umpires to check Joe Musgrove for any foreign substances. Throughout the game,
there was focus on Musgrove's ears and how his spin rates and velocity were higher
than average. In addition, Musgrove was giving the Mets a hard time and was 
keeping them scoreless. 

When the umpires checked Musgrove, they rubbed his ears and found no foreign
substances. Musgrove continued to pitch and continued to dominate on the mound.
In this report, I will investigate whether it was possible Musgrove was 
cheating and compare this game to his regular season performance.

## Processing the Data


This data was gathered from a Kaggle dataset that was modified from Baseball 
Savant. This is a public source and is in association with the MLB for statcast
and other baseball statistics. They gather their data directly from 
MLB Advanced Media, L.P.


```{r, include = FALSE}
library(tidyverse)
library(lubridate)
```


Changed the format of the game_date from string to Date.
```{r loading the data, echo = F }
dat <- read_csv("C:/Users/kashi/OneDrive/Desktop/joe_musgrove/MLB_Musgrove-Joe_Reg-Post-Season_Pitch-Data.csv")
full_dat <- read_csv("C:/Users/kashi/OneDrive/Desktop/joe_musgrove/savant_data.csv")
dat <- dat %>%
  mutate(game_date=as.Date(game_date, "%m/%d/%Y"))
head(dat)

```


## Analysis


### Mets Game VS. 2022 Regular Season


I will investigate the average spin rate for each pitch type including: slider,
curveball, 4-seam fastball, changeup, sinker, and cutter. I will compare the 
spin rates with the individual Mets game and the entire 2022 season.
```{r mets game vs Regular Season, results='hide'}
mets <- dat%>%
  filter(game_date == '2022-10-09')%>%
  arrange(at_bat_number, strikes, at_bat_outcome)
head(mets)

mets_avg_spin <- mets%>%
  group_by(pitch_name)%>%
  summarise(avg_spin_rate = mean(release_spin_rate))%>%
  mutate(season_type = 'Mets Game')

reg_spin_rate <- dat%>%
  filter(season_type == 'Regular Season')%>%
  group_by(pitch_name)%>%
  summarise(avg_spin_rate = mean(release_spin_rate))%>%
  mutate(season_type = 'Regular Season')

mets_vs_reg_spinrate <- rbind(mets_avg_spin, reg_spin_rate)
```


The following plot shows an increase in spin rates for all of Musgrove's pitches.
On average, Musgrove had a higher spin rate than his season's average. This
will obviously open some eyes to check for anything fishy or in other words, 
sticky.

We see that Slider had a significant jump in RPM's during the Mets game. Sliders
and Curveballs are called breaking balls and have the most movement, therefore, 
they will have the highest spin rates. We see this and both pitches had an
increase in spin rates during the Mets game.

```{r mets plot, echo = F}
ggplot(data = mets_vs_reg_spinrate) +
  geom_point(aes(x = pitch_name, y = avg_spin_rate, shape = season_type, color =
                   season_type), size = 4) + 
  scale_y_continuous(name="Spin Rate", limits=c(1900, 3000))+
 annotate('text', x=6, y=2850, label='+249.19 RPM', color = 'red') +
  annotate('text', x=2.3, y=2050, label='+171.15 RPM') +
  annotate('text', x=1.6, y=2620, label='+106.71 RPM') +
  annotate('text', x=3.6, y=2800, label='+176.64 RPM') +
  annotate('text', x=4.6, y=2640, label='+116.67 RPM') +
  annotate('text', x=5.6, y=2500, label='+105.55 RPM') +
  labs(title = "Average Spin Rate", subtitle = "Mets Wild Card Game Vs. Regular Season",
       caption = "Data source: Baseball Savant", color ="", shape = "",
       x = "Pitch Type")
```

### Regular Season VS Post Season


I will compare the spin rates between the regular season and the post season.
```{r Regular Season vs Post Season, include = F}
avg_spin <- dat%>%
  group_by(season_type, pitch_name)%>%
  summarise(avg_spin_rate = mean(release_spin_rate), .groups = 'drop')
```

The following plots shows that the average spin rate during the post season was
higher than the regular season. This does take in account of the Mets game,
however, we found that the averages in the post season were still higher.

This shows that it was not only the Mets game, but the other playoff games 
displayed a higher spin rate for all pitches.

```{r reg season vs post season, echo = F}
ggplot(data = avg_spin)+
  geom_point(aes(x = pitch_name, y = avg_spin_rate, shape = season_type, color = season_type), size = 4)+
  annotate('text', x=6, y=2850, label='+192.60 RPM', color = 'red') +
  annotate('text', x=2.3, y=2020, label='+101.94 RPM') +
  annotate('text', x=1.6, y=2620, label='+90.31 RPM') +
  annotate('text', x=3.6, y=2800, label='+116.26 RPM') +
  annotate('text', x=4.6, y=2640, label='+83.30 RPM') +
  annotate('text', x=5.6, y=2500, label='+79.35 RPM') +
  labs(title = "Average Spin Rate", subtitle = "Post Season Vs. Regular Season",
       caption = "Data source: Baseball Savant", color ="", shape = "",
       x = "Pitch Type")


post_without_mets <- dat%>%
  group_by(pitch_name)%>%
  filter(game_date != "2022-10-09" & season_type == 'Post Season')%>%
  summarise(avg_spin_rate = mean(release_spin_rate))


ggplot(data = post_without_mets, aes(x = pitch_name, y = avg_spin_rate, color = "pink"))+
  geom_point()+
  geom_point( data = reg_spin_rate, aes(x = pitch_name, y = avg_spin_rate, color = "blue"))+
  annotate('text', x=6, y=2800, label='+165.11 RPM', color = 'red') +
  annotate('text', x=2.3, y=2020, label='+80.60 RPM') +
  annotate('text', x=1.6, y=2620, label='+83.61 RPM') +
  annotate('text', x=3.6, y=2800, label='+101.39 RPM') +
  annotate('text', x=4.6, y=2640, label='+60.21 RPM') +
  annotate('text', x=5.6, y=2500, label='+60.35 RPM') +
  labs(title = "Average Spin Rate", subtitle = "Post Season (w/o Mets Game) Vs. Regular Season",
       caption = "Data source: Baseball Savant", color ="", shape = "",
       x = "Pitch Type", y = "Spin Rate")+
  scale_colour_manual(values =c('pink'='pink','blue'='blue'), labels = c('Post Season','Regular Season'))
```


### More Visualizations


#### Trends


I created a line graph showing over time the average spin rates for each pitch.
We see that there are fluctuations in the spin rates, however, upon closer
examination we see a large positive trend at the end of September, where the 
playoffs begin.

```{r check the yearly data of each pitch and its average spin rate, echo = F}
reg_spin <- dat%>%
  group_by(pitch_name, game_date)%>%
  summarise(avg_spin_rate = mean(release_spin_rate), .groups = 'drop')

ggplot(data = reg_spin)+
  geom_line(aes(x = game_date, y = avg_spin_rate, color = pitch_name))+
  scale_x_date(date_labels="%b",date_breaks  ="1 month")+
  geom_hline(yintercept = 2559.32, linetype = 3)+
  geom_hline(yintercept = 1971.61, linetype = 3)+
  geom_hline(yintercept = 2721.77, linetype = 3)+
  geom_hline(yintercept = 2581.22, linetype = 3)+
  geom_hline(yintercept = 2435.65, linetype = 3)+
  geom_hline(yintercept = 2714.75, linetype = 3)+
  labs(title = "Average Spin Rates for 2022 Season", caption = "Data Source: baseball Savant",
       subtitle = "Every game Joe Musgrove started",
       x = "Date", y = "Spin Rate")
  
post_spin <- dat %>%
  filter(game_date >= '2022-09-01')%>%
  group_by(pitch_name, game_date)%>%
  summarise(avg_spin_rate = mean(release_spin_rate), .groups = 'drop')

ggplot(data = post_spin)+
  geom_line(aes(x = game_date, y = avg_spin_rate, color = pitch_name))+
  scale_x_date(date_labels="%m/%d",date_breaks  ="1 week")+
  labs(title = "Average Spin Rates for 2022 Season", caption = "Data Source: baseball Savant",
       subtitle = "September through October Games",
       x = "Date", y = "Spin Rate") +
  geom_vline(xintercept = as.numeric(post_spin$game_date[7]), linetype = 4)
```
#### Distribution of Pitches and Spin Rates


I used boxplots and histograms to show how Musgrove's pitches were distributed by 
spin rate. I was able to see the difference in distributions between the 
regular season and the post season as well as the regular season and the Mets 
game. I found that the regular season distribution looks to have normality
and this will be useful for a statistical analysis later.

```{r distribution of pitches, echo = F}
ggplot(dat)+
  geom_histogram(aes(x = release_spin_rate, color = season_type), fill = "white", alpha = 0.2, binwidth= 30, position = "identity") +
  facet_wrap(~pitch_name, nrow = 3)+
  labs(title = "Distribution of Pitches", x = "Spin Rate", color="", caption = "Data Source:Baseball Savant")
```

```{r boxplots, echo = F}
ggplot(data = dat)+
  geom_boxplot(aes(x = season_type, y = release_spin_rate))+
  facet_wrap(~pitch_name)+
  labs(title = "Distribution of Spin Rates", caption = "Data source:Baseball Savant", x = "Season Type", y= "Spin Rate")
  

dat%>%
  filter(game_date <= '2022-10-09')%>%
  ggplot() +
geom_boxplot(aes(x = season_type, y = release_spin_rate))+
  facet_wrap(~pitch_name)+
  labs(title = "Distribution of Spin Rates", caption = "Data source:Baseball Savant", x = "Season Type", y= "Spin Rate", subtitle = "Mets Game VS Regular Season") +
  scale_x_discrete(labels=c("Post Season" = "Mets Game"))
```


#### Statistical Analysis of Average Spin Rates


We saw that there is a significant difference in average spin rates between the 
regular season and post season. To give more evidence, we will run a t.test to
compare the averages of the Mets game with the regular season and the post season
with the regular season.

I ran a log transformation to consider the assumption of normality for a smaller
sample size for the post season and the Mets game. A wilcox test was also done to
account for these assumptions being ignored.


```{r statistical analysis, echo = F}
postseason <- dat%>%
  filter(season_type=='Post Season')
regular_season <- dat%>%
  filter(season_type == 'Regular Season')

postseason_curve <- postseason%>%
  filter(pitch_name == 'Curveball')

mets_curve <- mets%>%
  filter(pitch_name == 'Curveball')

regular_season_curve <- regular_season%>%
  filter(pitch_name == 'Curveball')
```

Below we find that the post season curveball average spin rate is statistically
greater than the regular season:

```{r, echo = F}
t.test(postseason_curve$release_spin_rate, mu = mean(regular_season_curve$release_spin_rate),
       alternative = "greater")

t.test(log(postseason_curve$release_spin_rate), mu = log(mean(regular_season_curve$release_spin_rate)),
       alternative = "greater")

wilcox.test(postseason_curve$release_spin_rate, mu = mean(regular_season_curve$release_spin_rate),
       alternative = "greater")

```


Below we find that the post season slider average spin rate is statistically
greater than the regular season:


```{r, echo = F}
postseason_slider <- postseason%>%
  filter(pitch_name == 'Slider')

mets_slider <- mets%>%
  filter(pitch_name == 'Slider')

regular_season_slider <- regular_season%>%
  filter(pitch_name == 'Slider')

t.test(postseason_slider$release_spin_rate, mu = mean(regular_season_slider$release_spin_rate),
       alternative = "greater")

t.test(log(postseason_slider$release_spin_rate), mu = log(mean(regular_season_slider$release_spin_rate)),
       alternative = "greater")

wilcox.test(postseason_slider$release_spin_rate, mu = mean(regular_season_slider$release_spin_rate),
       alternative = "greater")
```


Below we find that the mets game curveball average spin rate is statistically
greater than the regular season:

```{r, echo = F}

t.test(mets_curve$release_spin_rate, mu = 
       mean(regular_season_curve$release_spin_rate),
       alternative = "greater")

t.test(log(mets_curve$release_spin_rate), mu = log(mean(regular_season_curve$release_spin_rate)),
       alternative = "greater")

wilcox.test(mets_curve$release_spin_rate, mu = mean(regular_season_curve$release_spin_rate),
       alternative = "greater")
```

Below we find that the mets game slider average spin rate is statistically
greater than the regular season:

```{r, echo = F}
t.test(mets_slider$release_spin_rate, mu = mean(regular_season_slider$release_spin_rate),
       alternative = "greater")

t.test(log(mets_slider$release_spin_rate), mu = log(mean(regular_season_slider$release_spin_rate)),
       alternative = "greater")

wilcox.test(mets_slider$release_spin_rate, mu = mean(regular_season_slider$release_spin_rate),
       alternative = "greater")


```


Below we find that the postseason average spin rate is statistically
greater than the regular season:


```{r, echo = F}
t.test(mets$release_spin_rate, mu = mean(regular_season$release_spin_rate),
       alternative = "greater")

t.test(log(mets$release_spin_rate), mu = log(mean(regular_season$release_spin_rate)),
       alternative = "greater")

wilcox.test(mets$release_spin_rate, mu = mean(regular_season$release_spin_rate),
       alternative = "greater")
```


Overall, we found in all cases that the average spin rates during the post season
and the mets game were statistically greater than the regular season average.


## Conclusion

Based on the recurring pattern and analysis of higher spin rates during the Mets
game compared to the regular season, we can believe that Joe Musgrove could have
been using a foreign substance to increase his pitching spin rate. The statistics
and the visuals show these increases can infer a substance use.

However, we know that the umpires did a substance check on Joe Musgrove and they
found nothing suspicious. Is this just complacency by the umpires? Or did they
miss a spot? We will never know unless Musgrove confesses to tell the truth. All
we know is that Musgrove's spin rates were higher than average. Therefore, he
was either cheating or he was just having a great game as an ace. 

There are other factors that can attribute to this higher spin rate. For example,
there were other MLB players indicating that players use "Red Hot" on their ears
to get them psyched up and pitch or perform better. Another factor is that it was
the playoffs and these are the most important games for players. Musgrove must 
have been focused and more psyched to do his best, therefore, we see higher numbers
during the playoffs.

An argument that goes against the use of foreign substances is that there was not
an enormous change in spin rate in Musgrove's game compared to recent studies on
the use of foreign substances. Another [study](https://proplayai.com/a-sticky-situation/) found that spin rates increased from 1600 rpm to 2200 rpm, a 600 rpm increase. 
This is a huge difference compared to Musgrove's increase of 250 rpm. Therefore,
it could be the case that there was no foreign substance.

In conclusion, it is possible that Joe Musgrove was cheating based on the higher
numbers across the board. However, we must consider the situation and other factors.
Cheating is never good for the game, but it is always great to see a pitcher
perform at their best.
